Wood density and the growth-dependent & independent mortality

### 2015-09-02 ms.Rmd
Updated manuscript. Still need to fix up cross validation section

where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25. We modelled $R_i$ as a random realisation from a lognormal distribution centered on zero and a standard deviation of 1, conveying our prior expectation that annual increases in dbh would be between 0.13 and 7.4 cm/yr. We used a lognormal as it ensures dbh growth cannot be negative.

We calculated basal area growth rate using: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.

We then calculated the mean posterior growth rate estimate for each individual and used these as our estimates of true dbh growth for the subsequent mortality analysis.

@@

We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1997vy] to include a growth-independent baseline hazard, such that the functional form becomes
\begin{equation} \label{eq:lambda}
\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
\end{equation}

@@

To incorporate varability in species mortality rates, we add allowed $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to encapsulate a species growth strategy (REFs). In addition to this, we also allowed the three parameters to vary by species ID and census period, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species, census and individual levels [@Gelman:2007te; @Kery:2010tp].

@@

To account for species strategy and partition variance into explained and unexplained at the observation, species and sampling period level we modelled $\alpha$, $\beta$ and $\gamma$ via an equation of the form:
\begin{equation} \label{eq:alpha}
\alpha_s = \alpha_{0,s} \, \alpha_{1,t} \, \rho_s^{\alpha_{2}}.
\end{equation}
Here $\alpha_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs, $\alpha_{1,t}$ represents the sample-level random effect for which time $t$ an individual was sampled, and $\alpha_{2}$ represents the effect of standardised wood density $\rho_s$ on $\alpha_s$. The form of eq \ref{eq:alpha} ensures that parameters remain positive. On a log scale equation \ref{eq:alpha} becomes an additive model,$\log \alpha_s = \log \alpha_{0,s} + \log \alpha_{1,t} + \alpha_{2} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis. We standardised wood density via the transformation: $\rho_s = \rho/0.6$, an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$. This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].

The species and census-level random effects $\alpha_{0,s}, \alpha_{1,t}$ were assumed to follow a log-normal distribution:
\begin{equation} \label{eq:alpha0_alpha1}
\alpha_i \sim \text{log-normal}\left(\mu_{\log \alpha_i}, \sigma_{\log \alpha_i}\right),\\
\end{equation}
where $\mu_{\log \alpha_i},  \sigma_{\log \alpha_i}$ are the mean and standard deviation of log-transformed variables. We estimated $\mu_{\log \alpha_{0,s}}$, the average species effect at zero growth using the following informative prior:

\begin{equation} \label{eq:alpha0}
\mu_{\log \alpha_{0,s}} \sim \text{normal}\left(-0.61, 0.57\right).\\
\end{equation}

This distribution was derived from data presented by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America. Note that these data are also used to define the growth-mortality relationship in SORTIE-ND a spatially explicit forest dynamic model (REF). We converted these probabilities to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e. -0.61) associated standard error (i.e. 0.). Based on this prior, we assumed that the average probability of death for an average species was 42%/yr with 95% of species effects occuring between 16% and 81%.

@@

Similarly, we used prior information from Korning and Balslev et al (1994) to estimate the average species effect on $\gamma$, the growth-independent hazard rate, such that:

\begin{equation} \label{eq:gamma0}
\mu_{\log \gamma_{0,s}} \sim \text{normal}\left(-3.86, 0.2\right).\\
\end{equation}

This distribution was estimated the same way as in $\mu_{\log \alpha_{0,s}}$. That is we converted annual mortality probabilities of 17 tropical lowland tree species in Amazonian Ecuador that encompass the range of forest structure (i.e. understory species, small stemmed midcanopy species, large stemmed midcanopy species and canopy species) to instantaneous hazard rate, and then log transformed these and calculated the mean (i.e. -3.86) and standard error (i.e. 0.2) These annual mortality estimates were averaged over size and growth rates, and as such are likely to be an overestimate of the true value of $\gamma$. However, based on this prior, we assume that the average growth-independent species mortality will be 2.1%/yr with 95% of species growth-independent probabilities of occuring between 1.4% and 3.1%.

@@

As we are fitting our model to two growth measures, we used weak priors to define the common log-normal distribution for the species effects on $\beta$:

\begin{equation} \label{eq:beta_0,s}
\mu_{\log \beta_{0,s}} \sim \text{lognormal}\left(0, 5\right).\\
\end{equation}

@@

The census random effects, $\alpha_1$, $\beta_1$ and $\gamma_1$ were all sampled from lognormal distributions centered on zero with a standard deviations estimated from the data. For example:

\begin{equation} \label{eq:a1b1c1,s}
\alpha_{1,t} \sim \text{lognormal}\left(0, \sigma_{a1}\right).\\
\end{equation}

The effect of wood density on all three parameters (i.e. $\alpha_2$, $\beta_2$ and $\gamma_2$) were given normal priors centered on zero and a standard deviation of 5:

@@

Standard deviations associated with both species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$) and those associated with census random effects (i.e. $\sigma_{a1}$, $\sigma_{b1}$, $\sigma_{c1}$), were all sampled from half cauchy priors with centered on zero with a scale of 2.5.

\begin{equation} \label{eq:a1b1c1,s}
\sigma_{a0},\sigma_{b0},\sigma_{c0},\sigma_{a1},\sigma_{b1},\sigma_{c1} \sim \text{cauchy+}\left(0, 2.5\right).\\
\end{equation}

## STAN MODEL
```{r, eval=FALSE}
data {
  int<lower=1> n_obs;
  int<lower=1> n_spp;
  int<lower=1> spp[n_obs];
  int<lower=1> n_census;
  int<lower=1> census[n_obs];
  int<lower=0, upper=1> y[n_obs];
  vector[n_obs] census_length;
  vector[n_obs] growth_dt;
  vector[n_spp] rho_c;

  // Held out data
  int<lower=1> n_obs_heldout;
  int<lower=1> n_spp_heldout;
  int<lower=1> spp_heldout[n_obs_heldout];
  int<lower=1> census_heldout[n_obs_heldout];
  int<lower=0, upper=1> y_heldout[n_obs_heldout];
  vector[n_obs_heldout] census_length_heldout;
  vector[n_obs_heldout] growth_dt_heldout;
  vector[n_spp_heldout] rho_c_heldout;
}

  real raw_log_a1[n_census];
  real<lower=0> sigma_log_a1;

  real raw_log_b1[n_census];
  real<lower=0> sigma_log_b1;

  real raw_log_c1[n_census];
  real<lower=0> sigma_log_c1;

  real a2;
  real b2;
  real c2;
}

  // Declaring mortality parameters
  real alpha;
  real a0[n_spp];
  real a1[n_census];

  real beta;
  real b0[n_spp];
  real b1[n_census];

  real gamma;
  real c0[n_spp];
  real c1[n_census];

  // Calculating census period random effects
  for (t in 1:n_census) {
    a1[t] <- exp(raw_log_a1[t] * sigma_log_a1); // e.g. implies lognormal(0, sigma_log_a1)
    b1[t] <- exp(raw_log_b1[t] * sigma_log_b1);
    c1[t] <- exp(raw_log_c1[t] * sigma_log_c1);
  }

  for (i in 1:n_obs) {
    // Calculating mortality parameters
    alpha <- a0[spp[i]] * a1[census[i]] * pow(rho_c[spp[i]], a2);
    beta <- b0[spp[i]] * b1[census[i]] * pow(rho_c[spp[i]], b2);
    gamma <- c0[spp[i]] * c1[census[i]] * pow(rho_c[spp[i]], c2);

  //Mortality model priors
  raw_log_a0 ~ normal(0,1);
  mu_log_a0 ~ normal(-0.87, 0.75);
  sigma_log_a0 ~ cauchy(0, 2.5);

  raw_log_c0 ~ normal(0, 1);
  mu_log_c0 ~ normal(-4.43, 1);
  sigma_log_c0 ~ cauchy(0, 2.5);

  raw_log_a1 ~ normal(0,1);
  sigma_log_a1 ~ cauchy(0, 2.5);

  raw_log_b1 ~ normal(0, 1);
  sigma_log_b1 ~ cauchy(0, 2.5);

  raw_log_c1 ~ normal(0, 1);
  sigma_log_c1 ~ cauchy(0, 2.5);

  a2 ~ normal(0,5);
  b2 ~ normal(0,5);
  c2 ~ normal(0,5);
}

  // Declaring fitted parameters
  real a0[n_spp];
  real a1[n_census];

  real b0[n_spp];
  real b1[n_census];

  real c0[n_spp];
  real c1[n_census];

  real alpha_fit;
  real beta_fit;
  real gamma_fit;
  real cumulative_hazard_fit;
  real log_lik_fit;
  real sum_log_lik_fit;

  // Declaring heldout parameters
  real alpha_heldout;
  real beta_heldout;
  real gamma_heldout;
  real cumulative_hazard_heldout;
  real log_lik_heldout[n_obs_heldout];
  real sum_log_lik_heldout;

  // recalulate species random effects
  for (t in 1:n_census) {
    a1[t] <- exp(raw_log_a1[t] * sigma_log_a1);
    b1[t] <- exp(raw_log_b1[t] * sigma_log_b1);
    c1[t] <- exp(raw_log_c1[t] * sigma_log_c1);
  }

  // log likelihood for fitted model
  for (i in 1:n_obs) {
    alpha_fit <- a0[spp[i]] * a1[census[i]] * pow(rho_c[spp[i]], a2);
    beta_fit <- b0[spp[i]] * b1[census[i]] * pow(rho_c[spp[i]], b2);
    gamma_fit <- c0[spp[i]] * c1[census[i]] * pow(rho_c[spp[i]], c2);

  // log likelihood for held out data
  for (j in 1:n_obs_heldout) {
    alpha_heldout <- a0[spp_heldout[j]] * a1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], a2);
    beta_heldout <- b0[spp_heldout[j]] * b1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], b2);
    gamma_heldout <- c0[spp_heldout[j]] * c1[census_heldout[j]] * pow(rho_c_heldout[spp_heldout[j]], c2);

@@

We will implement *K-fold cross validation*.

According to [Vehtari & Gelman 2014](http://www.stat.columbia.edu/~gelman/research/unpublished/waic_stan.pdf):

@@

>>>>

where, $R_i$ represents the true annual increase in dbh (cm) and $T_i$ represents the number of days between measurements divided by 365.25. We modelled $R_i$ as a random realisation from a half normal distribution centered on zero and a standard deviation of 5, conveying our prior expectation that annual increases in dbh would be between 0 and 10 cm/yr. We used a half normal as it ensures dbh growth cannot be negative.

\begin{equation} \label{eq:obs dbh increment}
R_i \sim \textrm{N+}(0, 5). \end{equation}.

For computational efficiency, we used the posterior means of $\dbh_t-1$, $dbh_t$ to calculate:

dbh growth rate:  $\frac{\widehat{dbh_{i,t}} - \widehat{dbh_{i,t-1}}}{T_i}$ 
basal area growth rate: $\frac{(\pi \, 0.25 \,\widehat{dbh_{i,t}}) - (\pi \, 0.25 \, \widehat{dbh_{i,t-1}})}{T_i}$.

These estimates of individual post growth were then used in the subsequent mortality analysis.

@@

We extend a common functional form used to model the growth-mortality relationship [e.g. @Kobe:1995] to include a growth-independent baseline hazard, such that the functional form becomes
\begin{equation} \label{eq:lambda}
\lambda = \overbrace{\alpha \, \exp\left(-\beta \, G \right)}^{\text{Growth dependent}} + \overbrace{\gamma}^{\text{Growth independent}}
\end{equation}

@@

To incorporate varability in species mortality rates, we add allowed $\alpha$, $\beta$ and $\gamma$ to vary by wood density - a species trait thought to encapsulate a species growth strategy (REFs). In addition to this, we also allowed the three parameters to vary by species ID, by doing this, we effectively allow the model to partition variation explained by wood density from the residual variation explained by species and individual levels [@Gelman:2007te; @Kery:2010tp].

@@

To account for species strategy and partition variance into explained and unexplained at the observation and species level we modelled $\alpha$, $\beta$ and $\gamma$ via the following submodels:
\begin{equation} \label{eq:submodels}
\alpha_s = \alpha_{0,s} \, \rho_s^{\alpha_{1}},
\beta_s = \beta_{0,s} \, \rho_s^{\alpha_{1}},
\gamma_s = \gamma_{0,s} \, \rho_s^{\alpha_{1}}.
\end{equation}
Here $\alpha_{0,s}$,$\beta_{0,s}$,$\gamma_{0,s}$ represents the species-level random effect of species $s$ within which individual $i$ belongs and $\alpha_{1}$,$\beta_{1}$,$\gamma_{1}$ represents the effect of standardised wood density $\rho_s$ on $\alpha$,$\beta$ and $\gamma$, respectively. The form of submodels in eq \ref{eq:submodels} ensures that parameters remain positive. On a log scale these submodels \ref{eq:submodels} become additive (e.g. $\log \alpha_s = \log \alpha_{0,s} + \alpha_{1} \, \log \rho_s$, equivalent to a classic proportional hazards model widely used in survival analysis. We standardised wood density via the transformation: $\rho_s = \rho/0.6$), an operation which is equivalent to centering a log-transformed variable at a density of 0.6 g cm${^-3}$. This was done for easier interpretation, allowing $\alpha_{1,t}$ to be interpreted as the hazard for a species with wood density of 0.6cm${^-3}$ (as opposed to species responses at $\rho$ = 0; an biologically unrealistic state for a woody plant) [@Gelman:2007te].

The species random effects on each parameter were assumed to be a random realistation from a log-normal distribution:
\begin{equation} \label{eq:RE_species}
\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \alpha_0}, \sigma_{\log \alpha_0}\right),\\
\beta_{0,s} \sim \text{log-normal}\left(\mu_{\log \beta_0}, \sigma_{\log \beta_0}\right),\\
\alpha_{0,s} \sim \text{log-normal}\left(\mu_{\log \gamma_0}, \sigma_{\log \gamma_0}\right),\\
\end{equation}

We estimated $\mu_{\log \alpha_0}$, the average log species effect at zero growth using the following informative prior:

\begin{equation} \label{eq:alpha0}
\mu_{\log \alpha_{0}} \sim \text{normal}\left(-0.61, 0.57\right).\\
\end{equation}

This distribution was derived from data published by Kobe et al 1995, who estimated the probability of dying at zero growth/2.5 years for juveniles of 10 hardwood species across north-eastern America. Note that these data are also used to define the growth-mortality relationship in SORTIE-ND a spatially explicit forest dynamic model (REF). We converted these probabilities/2.5 years to instantaneous hazard rates by using $\lambda = -\log(1 - Pr(Death))/2.5$, and then log transforming these values and calculating the mean (i.e. -0.61) and standard error (i.e. 0.57). Based on this prior, we assumed that the average probability of death was between 16 and 81% with the central tendency being 42%/yr.

@@

Similarly, we used prior information from Korning and Balslev et al (1994) to estimate, $\mu_{\log \gamma_0$, the average log species effect on $\gamma$, such that:

\begin{equation} \label{eq:gamma0}
\mu_{\log \gamma_0 \sim \text{normal}\left(-3.86, 0.2\right).\\
\end{equation}

This distribution was estimated the same way as in $\mu_{\log \alpha_0$. That is we converted annual mortality probabilities of 17 tropical lowland tree species in Amazonian Ecuador that encompass the range of forest structure (i.e. understory species, small stemmed midcanopy species, large stemmed midcanopy species and canopy species) to instantaneous hazard rate, and then log transformed these and calculated the mean (i.e. -3.86) and standard error (i.e. 0.2) These annual mortality estimates were averaged over size and growth rates, and as such are likely to be an overestimate of the true value of $\gamma$. However, based on this prior, we assume that the average growth-independent species mortality will be 2.1%/yr with 95% of species growth-independent probabilities of occuring between 1.4% and 3.1%.

@@

As we fitted our model to two growth measures (i.e. dbh growth and basal area growth), we used weak normal priors on $\mu_{\log \beta_0$, the average log species decay rate where:

\begin{equation} \label{eq:beta_0}
\mu_{\log \beta_0 \sim \text{normal}\left(0, 5\right).\\
\end{equation}

@@

The effect of wood density on all three parameters (i.e. $\alpha_1$, $\beta_1$ and $\gamma_1$) were given normal priors centered on zero and a standard deviation of 5:

@@

Standard deviations associated with species random effects (i.e. $\sigma_{a0}$, $\sigma_{b0}$, $\sigma_{c0}$), were all sampled from positive half cauchy priors with centered on zero with a scale of 2.5.

\begin{equation} \label{eq:a1b1c1,s}
\sigma_{a0},\sigma_{b0},\sigma_{c0} \sim \text{cauchy+}\left(0, 2.5\right).\\
\end{equation}

@@

@@

## STAN MODEL
```{r, eval=FALSE}
data {
  int<lower=1> n_obs;
  int<lower=1> n_spp;
  int<lower=1> spp[n_obs];
  int<lower=0, upper=1> y[n_obs];
  vector[n_obs] census_length;
  vector[n_obs] growth_dt;
  vector[n_spp] rho_c;

  // Held out data
  int<lower=1> n_obs_heldout;
  int<lower=1> n_spp_heldout;
  int<lower=1> spp_heldout[n_obs_heldout];
  int<lower=0, upper=1> y_heldout[n_obs_heldout];
  vector[n_obs_heldout] census_length_heldout;
  vector[n_obs_heldout] growth_dt_heldout;
  vector[n_spp_heldout] rho_c_heldout;
}

}

  // Declaring mortality parameters
  real alpha;
  real a0[n_spp];

  real beta;
  real b0[n_spp];

  real gamma;
  real c0[n_spp];

  for (i in 1:n_obs) {
    // Calculating mortality parameters
    alpha <- a0[spp[i]]; 
    beta <- b0[spp[i]]; 
    gamma <- c0[spp[i]];

  //Mortality model priors
  raw_log_a0 ~ normal(0,1);
  mu_log_a0 ~ normal(-0.61, 0.57);
  sigma_log_a0 ~ cauchy(0, 2.5);

  raw_log_c0 ~ normal(0, 1);
  mu_log_c0 ~ normal(-3.86, 0.2);
  sigma_log_c0 ~ cauchy(0, 2.5);

}

  // Declaring fitted parameters
  real a0[n_spp];
  real b0[n_spp];
  real c0[n_spp];

  real alpha_fit;
  real beta_fit;
  real gamma_fit;

  real cumulative_hazard_fit;
  real log_lik_fit;
  real sum_log_lik_fit;

  // Declaring heldout parameters
  real alpha_heldout;
  real beta_heldout;
  real gamma_heldout;

  real cumulative_hazard_heldout;
  real log_lik_heldout[n_obs_heldout];
  real sum_log_lik_heldout;

  // log likelihood for fitted model
  for (i in 1:n_obs) {
    alpha_fit <- a0[spp[i]]; 
    beta_fit <- b0[spp[i]]; 
    gamma_fit <- c0[spp[i]];

  // log likelihood for held out data
  for (j in 1:n_obs_heldout) {
    alpha_heldout <- a0[spp_heldout[j]]; 
    beta_heldout <- b0[spp_heldout[j]]; 
    gamma_heldout <- c0[spp_heldout[j]];
